Indicators for Ecosystem Condition in Norway

### 2023-06-08 trophic_levels-biomass_indicators.rmd
Render book with trophic level biomass

---
title: "TrophicLevelBiomassRatios"
author: "James D. M. Speed"
format: 
  html:
    embed-resources: TRUE
editor: visual
knitr:
  opts_knit:
    root.dir: "S://R projects//TrophicIntIndicators"
execute: 
  cache: true
---

@@

```{r}
#| label: Load packages
#| include: false
#| cache: TRUE

library(terra)
library(tidyterra)
library(ggplot2)
library(sf)
library(scales)
library(RColorBrewer)
library(tidyverse)

@@

#NPP
#Read in
npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")

@@

#Norwegian counties to help plotting
norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
#Simplify by county nr
norcounty<-norcounty_shp %>% 
  group_by(FylkeNr) %>%
  summarise(geometry = st_union(geometry))

@@

```{r}
#| label: PlotTotalHerbivoreBiomass
#| cache: TRUE
#Total herbivore biomass
#For some reason need to read the raster in again
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

@@

```{r}
#| label: Plot carnivore biomass
#| cache: TRUE
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")

@@

```{r}
#| label: Calculating expected herbivore biomass
#| cache: TRUE
npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]

expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  ggtitle("Expected herbivore biomass")+
  geom_spatraster(data=expected_herbivore_biomass)+facet_wrap(~lyr)+theme_bw()+
  scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))
writeRaster(expected_herbivore_biomass,"Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif",overwrite=TRUE)

@@

```{r}
#| label: DifferenceHerbivoreBiomass
#| cache: TRUE
expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

@@

writeRaster(diffherb,"Vertebrates/outputs/HerbivoreBiomassDiffExpected.tif",overwrite=TRUE)

@@

livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
totalcarnivores<-carnivorerast[[49:60]]

@@

writeRaster(expected_carnivore_biomass,"Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif",overwrite=TRUE)

writeRaster(diffcarn,"Vertebrates/outputs/CarnivoreBiomassDiffExpected.tif",overwrite=TRUE)

@@

From the above we see that Norway has had less carnivore biomass than expected based on the availabile herbivore biomass across the time period. Some counties and time periods (e.g. Hedmark today and Finnmark in 1907 had about the same carnivore biomass as predicted.

@@

#Forest
#AR50 type raster
artype50<-rast("T:\\vm\\alle\\GIS_data\\Norge\\Cartography\\N50\\AR50\\AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")
#artype50<-rast("Vertebrates/AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")

@@

#Load biomass rasters
npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")

#Project the AR50 to the NPP rasters
artype50_FR<-project(resample(artype50_F,npprast,method='near'),npprast)

@@

norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
 #Simplify by county nr
norcounty<-norcounty_shp %>% 
   group_by(FylkeNr) %>%
   summarise(geometry = st_union(geometry))

@@

 writeRaster(forest_diffcarn,"Vertebrates/outputs/forest_carnivores_diffexpected.tif",overwrite=TRUE)
 writeRaster(forest_diffherb,"Vertebrates/outputs/forest_herbivores_diffexpected.tif",overwrite=TRUE)
```

@@

forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")

norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
 #Simplify by county nr
norcounty<-norcounty_shp %>% 
   group_by(FylkeNr) %>%
   summarise(geometry = st_union(geometry))

>>>>

# Trophic level biomass ratios

<br />

_Author and date:_
James Speed

June 2023

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(terra)
library(tidyterra)
library(ggplot2)
library(sf)
library(scales)
library(RColorBrewer)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse)

# Set up conditional file paths
dir <- substr(getwd(), 1,2)

path <- ifelse(dir == "C:", 
               "R:/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/",
               "/data/R/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/")

pData <- ifelse(dir == "C:", 
               "P:/41201785_okologisk_tilstand_2022_2023/data/",
               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

@@

@@

#NPP
#Read in
npprast<-rast(paste0(pData,"/Trophic_levels/NPP.tiff"))
viltrast<-rast(paste0(pData,"/Trophic_levels/Vilt_RawBiomass.tiff"))
livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))
carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))

@@

#Norwegian counties to help plotting
norcounty_shp<-st_read(paste0(pData,"/Trophic_levels/"),"ViltdataCounty")
#Simplify by county nr
norcounty<-norcounty_shp %>% 
  group_by(FylkeNr) %>%
  summarise(geometry = st_union(geometry))

@@

```{r}
#| label: PlotTotalHerbivoreBiomass
#| cache: TRUE
#Total herbivore biomass
#For some reason need to read the raster in again
livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))

@@

```{r}
#| label: Plot carnivore biomass
#| cache: TRUE
carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))

@@

```{r}
#| label: Calculating expected herbivore biomass
#| cache: TRUE
# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
# npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]

expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  ggtitle("Expected herbivore biomass")+
  geom_spatraster(data=expected_herbivore_biomass)+facet_wrap(~lyr)+theme_bw()+
  scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))
#writeRaster(expected_herbivore_biomass,paste0(pData,"/Trophic_levels/expectedherbivorebiomass.tif"),overwrite=TRUE)

@@

```{r}
#| label: DifferenceHerbivoreBiomass
#| cache: TRUE
#expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
#livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

@@

#writeRaster(diffherb,paste0(pData,"/Trophic_levels/HerbivoreBiomassDiffExpected.tif"),overwrite=TRUE)

@@

#livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))

totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
#carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))
totalcarnivores<-carnivorerast[[49:60]]

@@

#writeRaster(expected_carnivore_biomass,paste0(pData,"/Trophic_levels/expectedcarnivorebiomass.tif"),overwrite=TRUE)

#writeRaster(diffcarn,paste0(pData,"/Trophic_levels/CarnivoreBiomassDiffExpected.tif"),overwrite=TRUE)

@@

From the above we see that Norway has had less carnivore biomass than expected based on the available herbivore biomass across the time period. Some counties and time periods (e.g. Hedmark today and Finnmark in 1907 had about the same carnivore biomass as predicted.

@@

#Forest
#AR50 type raster
artype50<-rast("/data/R/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/AR50_Norge.tif")

#artype50<-rast("Vertebrates/AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")

@@

#Load biomass rasters
# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
# viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
# livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
# carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
# expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
# expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")

#Project the AR50 to the NPP rasters

### NEED TO CHECK IF THIS IS THE CORRECT LAYER OR GET THE CORRECT ONE FROM JAMES

artype50_FR<-project(resample(artype50_F,npprast,method='near'),npprast)

@@

#norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
 #Simplify by county nr
norcounty<-norcounty_shp %>% 
   group_by(FylkeNr) %>%
   summarise(geometry = st_union(geometry))

@@

 #writeRaster(forest_diffcarn,paste0(pData,"/Trophic_levels/forest_carnivores_diffexpected.tif"),overwrite=TRUE)
 #writeRaster(forest_diffherb,paste0(pData,"/Trophic_levels/forest_herbivores_diffexpected.tif"),overwrite=TRUE)
```

@@

#forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
#forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")

#norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
 #Simplify by county nr
norcounty<-norcounty_shp %>% 
  group_by(FylkeNr) %>%
 summarise(geometry = st_union(geometry))
